

# Data table examples using public data

By Gene Leynes
Chicago Data Science

## Introduction

This code was used to demonstrate the features of the data.table package at the R User Group meetup held in Chicago on February 7th, 2013.

The data and project related to this exmaple comes from another meetup and code project.  I attended that meetup, and put together some simple visualizations for inspiration.  These visualizations were intended to be used as talking points for the project.  I repurposed this data to provide some `data.table` examples.

Meetup group: http://www.meetup.com/The-Chicago-Data-Visualization-Group/events/97690642/ <br>
Github project related to meetup: https://github.com/sc3/26thandcalifornia <br>
My related project: https://github.com/geneorama/26_and_California <br>

Exmples complied using `knitr`: <br>
Original meetup example: http://chicagodatascience.com/public/26th_and_California_example_visualizations.html <br>


## INITIALIZE
```{r}
## The following line is needed to compile to HTML, please ignore
if (basename(getwd()) != "data.table_demo") {setwd("..")}

rm(list=ls())
library(data.table)
library(ggplot2)
source("functions/ExtractIsoTime.R")
source("functions/wtf.R")
source("functions/NAsummary.R")
```

## READ CSV

This csv contains raw data collected from the Cook Count Sherrif's website.  It was downloaded using a web scraping utility maintained here: https://github.com/sc3/cookcountyjail


```{r}
## The following line is needed to compile to HTML, please ignore
if (basename(getwd()) != "data.table_demo") {setwd("..")}

rawdat = read.table(
	file = 'data/Database 2013-01-21 (8zQ4cW7T).csv', 
	sep=',', quote='"', flush=FALSE, 
	header=TRUE, nrows=-1, fill=FALSE, 
	stringsAsFactors=FALSE,
	na.strings=c('None', ''))

str(rawdat)
dat = as.data.table(rawdat)
str(dat)
```

## Convert booking and discharge dates to date / time objects

This relies on a simple function I wrote called `ExtractIsoTime` (which is in the functions folder).   The thing to notice here is that when you create new columns using `data.table` you use `:=` to denote assignment.

```{r}
## EXAMPLE FORMAT: 2012-12-30T20:57:19.616186
dat[ , booking_date := ExtractIsoTime(dat$booking_date)]
dat[ , discharge_date_earliest := ExtractIsoTime(dat$discharge_date_earliest)]
```


## Examples of grouping with data.table

```{r}

## Not directly related to data.table, but I like to see how many unique
## and how many missing values each column has.

NAsummary(dat)


## Summary by bail amount the old way:
## (Don't forget the useNA argument!!)
table(dat$bail_amount)
table(dat$bail_amount, useNA='ifany') # almost forgot

## Summary by bail amount Data Table
dat[ , .N, by = bail_amount]
dat[ , .N, keyby = bail_amount]

## Summary by race
dat[ , .N, by=race]
```


## Examples of subsetting with data.table

This is probably the most confusing thing at first

```{r}
## WRONG WAY:
dat[ , 3]
dat[ , 'race']

## RIGHT WAY:
dat[1:10 , race] 
dat[1:10 , 3, with = F]


## Indexing works differently 
dat[1]

## DF:
df = as.data.frame(dat)
# df[1]
df[1,]


```

## Examples of grouping and aggregating with data.table

This is very fast, and very useful for generating any kind of summary statistics.

```{r}

## Grouping is simple
dat[ , mean(age_at_booking), by=race]
dat[ , age_at_booking, by=race]
dat[ 
	i = TRUE, 
	j = list(
		mean = mean(age_at_booking),
		sd = sd(age_at_booking)
		), 
	by = race]
```

## Examples of grouping and aggregating with data.table

But look at what happens here!

Sometimes things happen that you may not expect. It's good, but possibly a surprise. By including `age_at_booking` without any aggregating function, it automatically expands the data.table result

```{r}
dat[ 
	i = TRUE, 
	j = list(
		mean = mean(age_at_booking),
		sd = sd(age_at_booking),
		age_at_booking
		), 
	by = race]
```

### Complex query example

```{r, tidy=FALSE}
mysummary = dat[
	i = !is.na(charges) &
		!is.na(booking_date) &
		!is.na(bail_amount),
	j = list(
		count = .N,
		coverage = diff(range(booking_date)),
		bailave = mean(bail_amount),
		bailsd = sd(bail_amount),
		bailmin = min(bail_amount),
		bailmax = max(bail_amount)
		),
	by = list(race,gender,age_at_booking)
	]
mysummary

## Uncomment to open summary as csv file, in Excel probably
# wtf(mysummary)
```

***
<br>
<br>
## NOTE: FOR THE REST OF THE EXAMPLES I'M GOING TO USE FEWER COLUMNS, JUST BECAUSE IT GETS VERY VERBOSE TO PRINT EVERYTHING
<br>
<br>
```{r}
datSmall = dat[,list(race, gender, charges_citation, housing_location)]
```


# J and CJ

Here are some examples of how to use (and not use) the Join (J) and Cross Join (CJ)

Personally, I rarely used these functions at first.  I found it easier to give up some performance for simplicity. 


```{r}

## Count of observations by race
datSmall[ , .N, by=race]

## Set the key to be "race"", and do some joins
setkey(datSmall, "race")

## Simple inner Join, new syntax (and what I normally use)
datSmall['W']
## Simple inner Join, old syntax
datSmall[J('W')]
## You can select more than one key
datSmall[J(c('W', 'WH'))]
##   This still works
datSmall[c('W', 'WH')]
```

## Examples with two keys: `race` and `gender`
```{r}
## Set the key
setkeyv(datSmall, c("race", "gender"))
```

Let's get records where race == "WH" and gender == "M"
There should be 1682 records of that sort based on this table:
```{r}
datSmall[,.N,keyby=list(race, gender)]
```

## Try using J and CJ on two keys

```{r}
## This is the first way that I would have guessed to 
## get results for "white" and "male", but it doesn't work:
datSmall[c('WH', 'M')]
## If I add the summary using .N, you can see that 
## it also selects Females (and returns 1 NA):
datSmall[c('WH', 'M')][,.N,list(race,gender)]

## This one works.
datSmall[data.table('WH', 'M')]
datSmall[data.table('WH', 'M')][,.N,list(race,gender)]

## But it's a little sloppy because you're actually
## making this data table, and then 
## joining it with the master data:
data.table('WH', 'M')

## This is the right way to do it.
datSmall[J('WH', 'M')]
datSmall[CJ('WH', 'M')]

## Notice that with two keys this no longer works:
datSmall[J('WH', 'W')]
## Notice that with two keys, but this does:
datSmall[J(c('WH', 'W'))]

```




